---
title: "External Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{External Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# External Data

It is important to have data to use for analysis. These data include bioenergetics parameters for the fish (or other species of interest) and water temperature. 

## Bioenergetics Parameters

The Fish Bioenergetics 4.0 software (Hanson et al. 1997, Deslauriers et al. 2017) provides a set of bioenergetics parameters for many species of fish from other sources, which also need to be cited. However, not all species are included in the software. One approach is to get a profile for the most closely related species to the one of interest. To do that, we use the phylogeny from the Open Tree of Life (Hinchliff et al. 2015) to find the closest species and then use the Fish Bioenergetics 4.0 software to get the parameters for that species. For example, suppose we wanted the parameters for the arctic grayling (*Thymallus arcticus*), to see what might happen if it were introduced to the eastern US. We can use the `fb4_closest_species` function to find the closest species in the Fish Bioenergetics 4.0 database:

```{r}
library(fishyenergy)
closest_species <- fb4_closest_species("Thymallus arcticus")
closest_species
```

We can then get information from the included dataset, `fb4_data`, to get the source for that species:

```{r}
print(fb4_data$Source[fb4_data$Sci_Name == closest_species])
```

We can then use the `bem_from_fb4` function to get the bioenergetics parameters for that species, if we supply the closest species name. We also have to include parameters that are not in the Fish Bioenergetics 4.0 software, such as the proportion of maximum consumption and other parameters. The values below are from Troia and Perkin (2022) for *Micropterus treculii* and are probably wrong for  *Thymallus arcticus*. 

```{r}
BEM <- bem_from_fb4(closest_species, CP=0.5, FA=0.1018428, UA=0.08743042, SDA=0.150819, ED=3983.646)
```

We might want to see how this species would grow in various places in Tennesssee, but we need information on the water temperature in those places. One source is the US Geological Survey (USGS) databases, which have that information. So let's find all the sites in Tennesssee that have water temperature data, using a wrapper for the `dataRetrieval` package:

```{r}
stations <- water_stations_find(state="TN")
knitr::kable(head(stations[,1:6], 5), caption="First 5 stations in Tennessee with water temperature data")
```

Let's see how they'd do in the Little River:

```{r}
little_river_row <- stations[stations$station_nm == "LITTLE RIVER ABOVE TOWNSEND, TN",]
little_river_id <- paste0("USGS-", little_river_row$site_no)
```

For this, we would need the water temperature for that site. Let's get it for 2023, padding a bit on each side so we can get a full year of data:

```{r}
raw_temps <- daily_temperature_raw(little_river_id, start_date="2022-11-01", end_date="2024-02-01")
```

This was measured manually, so there are not a lot of data points. Let's plot the data to see what it looks like:

```{r}
plot(raw_temps$ActivityStartDate, raw_temps$ResultMeasureValue, pch=19, xlab="Date", ylab="Temperature (C)", main="Little River Temperature Data", bty="n")
```

We need points for every day, so let's interpolate the data. By default, it will give us an error if there are a lot of missing data: by default, if it is over 80% missing. But we can manually change that threshold with the `minimum_fraction_missing` argument. Let's set it to 0.5, so it will give us an error if more than 50% of the data is missing. We can also focus on just 2023 by setting the `start_date` and `end_date` arguments:

```{r}
interpolated_temps <- daily_temperature_interpolate(raw_temps, minimum_fraction_missing=0.05, start_date="2023-01-01", end_date="2023-12-31")
```

We can see these on the original plot:

```{r}
plot(raw_temps$ActivityStartDate, raw_temps$ResultMeasureValue, pch=19, xlab="Date", ylab="Temperature (C)", main="Little River Temperature Data", bty="n")
points(interpolated_temps$date, interpolated_temps$temp, col="red", pch='.')
```

It's not a smoothed plot: we keep the actual temperatures the fish experienced, and just linearly interpolate between them. Now we can compute the weight gain for the fish at this site:

```{r}
station_result <- single_station_compute(T_vector=interpolated_temps$temp, BEM=BEM)
```

We can plot the weight gain over the year:

```{r}
plot(interpolated_temps$date, station_result$W2_cum, xlab="Date", ylab="Mass (g)", type='l', bty="n", main="Weight Gain in Little River")
```

We can look at every species in the database and see how they would have done in the Little River in TN. For this, we will make very, very bad assumptions: they all start with a weight of 6.382417 g, they all have the same joules per gram of wet mass for the species' prey, and they all have the same oxycalorific coefficient. If we knew these we could pass in a vector with one value per organism, including different values for different life stages if present. We also assume each is happy living in a freshwater temperate mountain stream -- very unlikely for a species like a lionfish, but we will throw it in. 

```{r}
all_organisms <- loop_species_across_station(interpolated_temps$temp)

all_organisms$SpeciesStage <- all_organisms$common

for (i in 1:nrow(all_organisms)) {
  if (nchar(all_organisms$life_stage[i]) > 0) {
	all_organisms$common[i] <- paste0(all_organisms$SpeciesStage[i], " (", all_organisms$life_stage[i], ")")
  }
}

somewhat_surviving_organisms <- all_organisms[!is.na(all_organisms$W2_cum),] 



weird_taxa <- unique(c(subset(somewhat_surviving_organisms, ProportionStartingWeight < 0.5 | ProportionStartingWeight > 25)$SpeciesStage, all_organisms$SpeciesStage[is.na(all_organisms$W2_cum)]))


feasible_organisms <- somewhat_surviving_organisms[!somewhat_surviving_organisms$SpeciesStage %in% weird_taxa,]

library(ggplot2)
library(ggrepel)
library(dplyr)

last_points <- feasible_organisms |>
  group_by(SpeciesStage) |>
  slice_max(order_by = julian, n = 1)


ggplot(feasible_organisms, aes(x=julian, y=ProportionStartingWeight, color=SpeciesStage)) +
  geom_line() + theme_minimal() +
  geom_text_repel(
    data = last_points,
    aes(label = SpeciesStage),
    hjust = 0, nudge_x = 5, size = 3, show.legend = FALSE, segment.color = 'gray', box.padding = 0.3, max.overlaps = Inf
  )  +
  labs(x="Day of year", y="Relative mass") + xlim(c(0, 500)) + theme(legend.position = "none")
```

We can also see what organism would not survive (would go to zero body mass) in the Little River in TN given our (bad) assumptions going into the analysis:

```{r}
dead_organisms <- sort(unique(somewhat_surviving_organisms$SpeciesStage[which(somewhat_surviving_organisms$W2_cum<0)]))
```

The organisms that would not survive are `r paste(dead_organisms, collapse=", ")`.

This is a very bad analysis, but it shows how to use the `fishyenergy` package to get bioenergetics parameters for a species, find water temperature data, and compute weight gain for that species at a site -- use it with your own better data!