---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Vignette objectives
- Learn fundamentals of bioenergetics modeling (BEM) by visualizing temperature- and mass-dependent physiological rates.
- Explore effects of temporal extrinsic parameters like temperature, the proportion of maximum consumption (CP), the activity multiplier (ACT), and prey energy density.
- Compare bioenergetics of a representative warm-water and cold-water species.
- Simulate daily bioenergetics and growth of a fish.
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

### Install and load fishyenergy
```{r setup}
remotes::install_github("bomeara/fishyenergy")
library(FishyEnergy)
```

### Temperature-dependent physiological rates
Let's look at temperature-dependent physiological rates of two species. Start by loading the intrinsic bioenergetics parameters of five example species. These parameters have been published by various authors since the 1980s and are available from [Fish Bioenergetics 4.0](http://fishbioenergetics.org/).
```{r}
options(width = 500)
data(parms_fb4)
```

Next, retain only one focal species, largemouth bass *Micropterus salmoides*, and look at these intrinsic parameters.
```{r}
options(width = 500)
parms.micsal <- parms_fb4[parms_fb4$genus_species == "micropterus_salmoides",]
parms.micsal
```
- The first column shows the Latin name in snake case.
- Other columns like lwA and lwB are the intercept and slope of the length-weight equation acquired from [FishBase](https://fishbase.org).
- Still other columns like CA and CB are the intercept and slope of the mass dependent consumption equation (a negative power law).
- There's also specific dynamic action (*i.e.*, the proportion of energy consumed allocated to the cost of digestion), predator energy density, and many other variables defined by Hanson *et al.* (1997).

### Plot largemouth bass curves
Let's visualize the effect of temperature on consumption, respiration, and various other physiological rates for largemouth bass. To do this, we use the **bem_curve** function which needs to be supplied with a few essential arguments:
- Set the mass to a reasonable amount (454 grams which is ~ 1 pound).
- We plot this curve over ecologically realistic temperature range of 5°C to 35°C.
- We specify the intrinsic parameters for largemouth bass and the corresponding consumption and respiration functions... 2 and 1, respectively.
```{r fig.height=4, fig.width=6}
bem_curve(M = 454,
          T = 5:35,
          CP = 1.0,
          ACT = 1.0,
          prey_ED = 4000,
          parms.intrinsic = parms.micsal,
          C_eq = 2,
          R_eq = 1)
```

Notice that the consumption curve increases exponentially to an optimum temperature of ~28°C and then declines at higher temperatures. This curve shape is characteristic of consumption function 2. In contrast, we see that the respiration curve increases exponentially over the full range of temperatures (*i.e.*, there is no optimum beyond which respiration declines). This curve shape is characteristic of respiration function 1.

Compare the above curves to those plotted when the activity multiplier is increased from 1.0 to 4.0, thus simulating a predator fish with 4 times the metabolic demand of the previous predator fish at rest. Pursuing prey, avoiding predators, and seeking reproductive opportunities typical ecological processes that increase metabolic demand beyond rest. Notice that the red respiration curve is higher and consequently the black surplus curve is lower.
```{r fig.height=4, fig.width=6}
bem_curve(M = 454,
          T = 5:35,
          CP = 1.0,
          ACT = 4.0,
          prey_ED = 4000,
          parms.intrinsic = parms.micsal,
          C_eq = 2,
          R_eq = 1)
```

We can also reduce the proportion of consumption from 1.0 to 0.5 to simulate ecological processes both bottom-up and top-down processes. Notice that the blue consumption curve is lower and consequently the black surplus curve is also lower.
```{r fig.height=4, fig.width=6}
bem_curve(M = 454,
          T = 5:35,
          CP = 0.5,
          ACT = 4.0,
          prey_ED = 4000,
          parms.intrinsic = parms.micsal,
          C_eq = 2,
          R_eq = 1)
```

Although we do not change prey energy density in this vignette, one can easily do this via the prey_ED argument to simulate different diets and/or prey quality.

### Plot brook trout curves
It is clear that largemouth bass perform well at higher temperatures which is why we know them to be members of the warmwater thermal guild. Let's visualize the effect of temperature on the same physiological rates for a putative coldwater species, the brook trout *Salvelinus fontinalis*. Notice that the consumption and respiration equations for brook trout are 2 and 2, respectively. Notice that we use the respiration2 equation for brook trout rather than the respiration1 equation. See Hanson *et al.* (1997) for details.
```{r fig.height=4, fig.width=6}
bem_curve(M = 454,
          T = 5:35,
          CP = 1.0,
          ACT = 1.0,
          prey_ED = 4000,
          parms.intrinsic = parms_fb4[parms_fb4$genus_species == "salvelinus_fontinalis",],
          C_eq = 2,
          R_eq = 2)
```

### Growing a fish
First, simulate (*i.e.*, create a fake) temperature time series with mean daily water temperature for every day of a calendar year. We do this using a sine wave function, setting mean annual temperature and annual amplitude to values that are realistic for a temperate waterbody that does not freeze-over during the winter.
```{r}
parm.tmean <- 16        # mean annual temperature
parm.ampli <- 13        # temperature amplitude
julian <- seq(1,365,1)
WT_mean <- round((parm.ampli * sin((2*(pi/365)) * julian - 2) + parm.tmean) * 10, digits = 0)
temps_fake <- data.frame(julian, WT_mean)
```

Plot the time series of mean daily water temperature.
```{r fig.height=4, fig.width=6}
library(ggplot2)
ggplot(NULL) + theme_classic() +
  labs(x = "Day of year", y = "Water temperature (°C)") +
  scale_x_continuous(expand = c(0,0), limits = c(0,366), breaks = seq(0, 360, 60)) + 
  scale_y_continuous(expand = c(0,0), limits = c(0,40), breaks = seq(0, 40, 10)) + 
  geom_line(aes(x = julian, y = WT_mean / 10), temps_fake, linewidth = 1)
```

### Temporally-varying parameters
We need to load some parameters that potentially vary temporally. The dataframe has 365 rows representing each day of the year. There are five noteworthy columns (*i.e.*, variables).
```{r}
options(width = 500)
data(parms_temporal_DEFAULT)
dim(parms_temporal_DEFAULT)
head(parms_temporal_DEFAULT)
```

- CP is the proportion of maximum consumption and decrease as prey quantity declines or as exploitative competition increases.
- ACT is the activity multiplier where a value of zero would indicate no metabolic costs beyond being at rest. Maintaining position in a flowing stream, avoiding predators, pursuing prey, and engaging in reproductive behavior increase ACT.
- The next two parameters (gsi_f and gsi_m) allow the user to model somatic growth of a reproductively mature fish that allocates some energy to gonadal tissue. Separate parameters are provided for females and males because the sexes invest different amounts of energy to their gonads--gsi_f and gsi_m, respectively. The default data assume a reproductively immature fish (*i.e.*, gsi = 0).
- Modifying prey energy density (prey_ED) over the year can simulate ontogenetic diet shifts or changes in prey quality. There are numerous published estimates of prey energy density for common prey items like invertebrates, fish, and so on (see Appendix B from Hanson *et al.* 1997).

### Simulating daily bioenergetics
Use the **bem_grow** function to simulate the daily bioenergetic budget of a largemouth bass. Start the fish at 454 grams which is ~1 pound. We need to load some parameters that potentially vary temporally.
```{r}
grow.micsal <- bem_grow(start_M2 = 454,
                        temperature = temps_fake,
                        parms.intrinsic = parms.micsal,
                        parms.temporal = parms_temporal_DEFAULT,
                        C_eq = 2,
                        R_eq = 1)
```

The output of bem_grow is quite extensive. There are 365 rows representing each day of the year and 43(!) variables. Let's have a look at those variables:
```{r}
options(width = 500)
head(grow.micsal)
```

- The first eight variables represent inputs that you're already familiar with like temperature, CP, ACT, gsi, and so on.
- The remaining variables are outputs including consumption (C), respiration (R), egestion (F), excretion (U), specific dynamic action (SDA), somatic mass (MS), and gonadal mass for females and males (MG).
- The "1" version of these variables are in units of energy (joules) and the "2" version of these variables are in units of mass (grams).
The "_ins" version of these variables represents instantaneous (*i.e.*, daily) values and the "_cum" represents cumulative values (*i.e.*, since January 1st).
- Finally, cumulative length (L_cum) and Fulton condition factor (K) are calculated from the length-weight parameters in parms_fb4. Survival (Surv) is assumed to start at 1 on January 1st and remains 1 if Fulton condition factor remains at or above 0.4 but switches to 0 if K falls below 0.4.

### Plotting daily energetics
Plot the time series of m2_cum. This is the cumulative mass of the predator fish in grams.
```{r fig.height=4, fig.width=6}
ggplot(NULL) + theme_classic() +
    labs(x = "Day of year", y = "Body mass (g)") +
    scale_x_continuous(expand = c(0,0), limits = c(0,365), breaks = seq(0,360,60)) +
    scale_y_continuous(expand = c(0,0), limits = c(0,4000), breaks = seq(0,4000,1000)) +
    geom_line(aes(x = julian, y = M2_cum), grow.micsal, linewidth = 1)
```

Notice that our largemouth bass does most of its growing during the warm summer months which makes sense for this warmwater species. What doesn't make sense is that this fish starts the year at 454 grams (~1 pound) and ends the year at a whopping 3,592 grams (7.9 pounds)!  This is because we unrealistically have assumed the predator fish has limitless food (CP = 1), sets up shop in one place and never moves (ACT = 1), and allocates no energy to reproduction (gsi_ = 0).

The **bem_corroborate** function provides a means of estimating difficult-to-measure parameters like CP and ACT using field- or lab-derived measurements of length-at-age to identify pairs of CP and ACT values that result in realistic end-of-year mass. See the [Corroboration vignette](https://bomeara.github.io/fishyenergy/articles/Corroboration.html) to learn this workflow.

To better visualize this summer growth, plot m2_ins. The horizontal red line is at y = 0, meaning that the predator fish has an energy deficit below this line (winter) and an energy surplus above this line (summer).
```{r fig.height=4, fig.width=6}
ggplot(NULL) + theme_classic() +
  labs(x = "Day of year", y = "Daily mass change (g)") +
  scale_x_continuous(expand = c(0,0), limits = c(0,366), breaks = seq(0, 360, 60)) +
  geom_hline(yintercept = 0, color = "red") +
  geom_line(aes(x = julian, y = M2_ins), grow.micsal, linewidth = 1)
```

Let's manually adjust CP and ACT to see how it affects growth. If dig into the literature, CP values are often around 0.6 and ACT values are often around 1.4. Change these values in the parms_temporal_DEFAULT dataframe, re-run the bem_grow function, and then re-plot growth.
```{r fig.height=4, fig.width=6}
parms_temporal_DEFAULT$CP <- 0.6
parms_temporal_DEFAULT$ACT <- 1.4

grow.micsal.real <- bem_grow(start_M2 = 454,
                             temperature = temps_fake,
                             parms.intrinsic = parms.micsal,
                             parms.temporal = parms_temporal_DEFAULT,
                             C_eq = 2,
                             R_eq = 1)
ggplot(NULL) + theme_classic() +
  labs(x = "Day of year", y = "Body mass (g)") +
  scale_x_continuous(expand = c(0,0), limits = c(0,366), breaks = seq(0, 360, 60)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,700), breaks = seq(0, 700, 100)) +
  geom_line(aes(x = julian, y = M2_cum), grow.micsal.real, linewidth = 1)
```

It seems reasonable that a 454 gram fish would grow to 548 grams in a year--that is, from 1.0 pounds to 1.2 pounds.

Finally, let's look at how mass, length, and condition change over time. Bioenergetics models are biologically realistic in that they simulate daily energy balance--that is, either deficits or surpluses. Mass can decrease or increase over time whereas length can increase with energy surpluses but cannot realistically decrease when there are energy deficits. The bem_grow function uses this axiom to simulate changes in Fulton condition factor over time as demonstrated below:

The previous plot shows that the mass of the predator fish starts at 454 grams on January 1st and decreases for 120 (until ~April). The predator fish then grows in both mass and length until day 290 (~October). Plot length (cm) over time to see how length does not change during periods of energy deficit and increases when there is an energy surplus.
```{r fig.height=4, fig.width=6}
ggplot(NULL) + theme_classic() +
  labs(x = "Day of year", y = "Total length (cm)") +
  scale_x_continuous(expand = c(0,0), limits = c(0,366), breaks = seq(0, 360, 60)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,50), breaks = seq(0, 50, 10)) +
  geom_line(aes(x = julian, y = L_cum), grow.micsal.real, linewidth = 1)
```

When mass decreases and length stays the same, Fulton condition should decrease.
```{r fig.height=4, fig.width=6}
ggplot(NULL) + theme_classic() +
  labs(x = "Day of year", y = "Fulton condition") +
  scale_x_continuous(expand = c(0,0), limits = c(0,366), breaks = seq(0, 360, 60)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,1.2), breaks = seq(0, 1.2, 0.2)) +
  geom_line(aes(x = julian, y = K), grow.micsal.real, linewidth = 1)
```

### Next steps
Now that you've had a crash course in the basics of bioenergetics and growing a fish, try the [Corroboration vignette](https://bomeara.github.io/fishyenergy/articles/Corroboration.html) where you will learn to grow fish that match the growth trajectories of real fish. Reach out to package developer [Matthew Troia](mailto:matthew.troia@utsa.edu) if you have questions about fishyenergy or suggestions to make it better!

### References
- Hanson, P. C., Johnson, T. B., Schindler, D. E., & Kitchell, J. F. (1997). Fish bioenergetics 3.0 for Windows.
