---
title: "Getting_Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting_Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fishyenergy)
```

First, let's load in our parameters for our model. For the vignette, we're loading it from a saved one in our package, but you can get one from your computer instead

```{r}
bem_raw <- read.csv(system.file("extdata","BEM_parameters_v01.csv", package="fishyenergy"))
chosen_row <- 1
bem_vector <- bem_raw[chosen_row,]
BEM <- new_BEM(bem_vector$CP, bem_vector$CA, bem_vector$CB, bem_vector$CTM, bem_vector$CTO, bem_vector$CQ, bem_vector$ACT, bem_vector$RA, bem_vector$RB, bem_vector$RTM, bem_vector$RTO, bem_vector$RQ, bem_vector$FA, bem_vector$UA, bem_vector$SDA, bem_vector$ED)
```

And then load in the temperatures for a year at a site:

```{r}
temp_raw <- read.csv(system.file("extdata","modeled_WT_HUC12100201.csv", package="fishyenergy"))
chosen_col <- 1
temp_vec <- temp_raw[,chosen_col]
```

Let's plot the temperature data to make sure they look like what we'd expect

```{r}
plot(x=as.Date(0:364, origin="2020-01-01"), y=temp_vec, xlab="Day", ylab="Temperature", type='l', bty="n")
```

Yay, looks good. Now let's compute the weight gain per day of our fishy:

```{r}
station_result <- compute_single_station(T_vector=temp_vec, BEM=BEM)
```

And look at how its weight changes over the year at that station:

```{r}
plot(x=as.Date(0:364, origin="2020-01-01"), y=station_result$W2_cum, xlab="Day", ylab="Mass (g)", type='l', bty="n")
```